---
title: "A_Short_Intro_to_norMmix"
output:
    html_document:
        fig_caption: yes
        theme: cerulean
        toc: yes
        toc_depth: 2
vignette: >
  %\VignetteIndexEntry{A_Short_Intro_to_norMmix}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  error = FALSE,
  tidy = FALSE,
  cache = FALSE
)
```

```{r setup}
library(norMmix)
set.seed(2020)
``` 

to dev and compile:
devtools::load_all()
rmarkdown::render("~/lib/norMmix/vignettes/A_Short_Intro_to_norMmix.Rmd")

## Ultra quick start

This section is for those who want a quick whiff of what the package can do.
For the proper start to this vignette, see next section.

A normal mixture model is a probability distribution constructed of normal distributions with weight coefficients

$$
f({\bf x}) = \sum_{k=1}^{K} \pi_k \Phi({\bf x};\mu_k, \Sigma_k) 
$$
$$
\text{ with } \sum \pi_k = 1, \quad \Phi(-; \mu_k, \Sigma_k) \text{ normal distribution with mean: } \mu_k \text{ and covariance matrix} \Sigma_k 
$$

`norMmix()` creates an object codifying such a distribution:

```{r}
w <- c(0.5, 0.3, 0.2)
mu <- matrix(1:6, 2, 3)
sig <- array(c(2,1,1,2,3,2,2,3,4,3,3,4), c(2,2,3))

nm <- norMmix(mu, Sigma=sig, weight=w)
plot(nm)
```

`rnorMmix()` generates data from the distribution:
```{r}
x <- rnorMmix(500, nm)
plot(nm)
points(x)
```

`norMmixMLE()` fits a distribution to data:
```{r}
ret <- norMmixMLE(x, 3, model="VVV", initFUN=claraInit)
plot(ret)
```

Voilà.

Now again but more thorough:


## Why this package exists:

This package was originally created for the purposes of a bachelor's thesis 
from author Nicolas Trutmann. The credit for the idea rests solely with Prof. 
Martin Mächler.

In brief: The current popular choice for the fitting of normal mixtures is the
EM-algorithm; in part because of it's proven convergence.
But, there are pathological cases where convergence slows down and, in most 
implementations of the EM-algorithm, break off.

This alternative, while not without flaws, is not hampered by this particular
shortcoming.
#### Rough outline:
* quickly contrast EM and alternative
* simple explanation of alt:
    + clara → m step → general optimiser
    + benefits of ldlt and loglik fctn as opposed to EM


* this document organised by major classes in package
    + norMmix: base class/base data model
    + norMmixMLE: return value of algo, holds results

This document is organised by the major classes in this package.
These are the 'norMmix', 'norMmixMLE' and (soon) 'manynorMmix' classes.

Of these, norMmix is the base class, that codifies a multivariate normal mixture,
with weights, means and covariance matrices.

Stacked on top of norMmix is norMmixMLE, which is the output of our main 
feature, the MLE implementation. It contains first and foremost a norMmix 
object, namely the result of the MLE. After that, it also has additional 
information about the specifics of the fitting, like number of parameters and 
sample size. 

## On the base class 'norMmix'

* how to make a norMmix model

It is easiest to introduce the tool by using the tool, so here is a quick tour:

```{r}
# suppose we wanted some mixture model, let
mu <- matrix(1:6, 2,3)      # 2x3 matrix -> 3 means of dimension 2
w <- c(0.5, 0.3, 0.2)       # needs to sum to 1
diags <- c(4, 3, 5)         # these will be the entries of the diagonal of the covariance matrices (see below)

nm <- norMmix(mu, Sigma=diags, weight=w)
print(nm)
str(nm)
```

A norMmix object is a list of 6: 
* a char denoting the model
* a matrix of means
* a vector of weights
* an array of covariance matrices
* an int: the number of components
* an int: the number of dimensions

The means and covariance matrices look like this:
```{r}
nm$mu
nm$Sigma
```
Compare that to mu and diags defined at the start of this section.

The norMmix() function serves as initializer for a norMmix object. While you 
could specify covariance matrices explicitly, norMmix() as a few nifty ways of 
constructing simpler matrices from smaller givens.
This happens according to the dimension of the given value for the Sigma argument:


0. for a single value `l` or NULL, norMmix() assumes all matrices to be diagonal with entries `l` or `1`, respectively.

1. for a vector `v`, `norMmix()` assumes all matrices to be diagonal with the i-th matrix having diagonal entries `v[i]`.

2. for a matrix `m`, `norMmix()` assumes all matrices to be diagonal with diagonal vector `m[,i]` (i.e. it goes by columns).

3. an array is assumed to be the covariance matrices, given explicitly.


### Issues with k or dim = 1

**IMPORTANT ISSUE**

* available functions for inspecting these
* what to do with them

## On 'norMmixMLE'

* the norMmixMLE function and its arguments
* Examples

Several development atavisms are still present in the code. 
These are exposed to the user in the package's namespace.
I will quickly show what these could be used for.


## On 'manyMLE' (as soon as we have reintegrated that)

## Details about the source code

* what else is in the code?
* grep -roP '\w+(?= <- function)' works well to get function names
    * grep -rn '<- function' also good
